---
title: "exo chap 2"
---

<style>
div.python pre { background-color: #e9f7f5
; }
</style>

<style>
div.r pre { background-color: #f8f8ff; }
</style>



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,error=TRUE,collapse = T)
``` 


<a href="3_Arrays_Vectors.html"> Go back to chapter 2</a>

<a href="3_Arrays_Vectors.html"> Go back to chapter 2</a>

# Exercice 1

The series, $1^{1} + 2^{2} + 3^{3} + ... + 10^{10} = 10405071317$.

Find the last ten digits of the series, $1^{1} + 2^{2} + 3^{3} + ... + 1000^{1000}$.

<div class = "r">
```{r}
# R

1000**1000

# To manage very large numbers you have to use gmp
library(gmp)

sum_n_power_last_digit <- function(n, last_digit){
  sum_series = sum(as.bigz(seq(1, n))**(seq(1, n)))
  sum_n = stringr::str_split(as.character(sum_series),'')[[1]]
  last_digits = sum_n[(length(sum_n)-last_digit+1):length(sum_n)]
  return(last_digits)
}

t = Sys.time()
sum_n_power_last_digit(1000,10)
Sys.time() - t

```

</div>

# Exercice 2


## Exercice 1 

If we list all the natural numbers below 10 that are multiples of 3 or 5, we get 3, 5, 6 and 9. The sum of these multiples is 23.

Find the sum of all the multiples of 3 or 5 below 10000.

<div class = "r">
```{r,collapse=T}
get_sum_multiples_below1 <- function(below,multiple_1,multiple_2){
  sum <- 0
  for(i in 1:(below-1)){
    if((i %% multiple_1 == 0)|(i %% multiple_2 == 0)){
      sum <-  sum + i 
    }
  }
  return(sum)
}

t <- Sys.time()
get_sum_multiples_below1(1000000,3,5)
Sys.time() - t



get_sum_multiples_below2 <- function(below,multiple_1,multiple_2){
  seq <- seq(1,(below-1))
  multiples <- sapply(seq,FUN = function(x){if((x %% multiple_1 == 0) | (x %% multiple_2 == 0)){x}else{0} })
  return(sum(multiples))
}

t <- Sys.time()
get_sum_multiples_below2(1000000,3,5)
Sys.time() - t



get_sum_multiples_below3 <- function(below,multiple_1,multiple_2){
  seq <- seq(1,(below-1))
  multiples <- seq[which((seq %% multiple_1 == 0) | (seq %% multiple_2 == 0))]
  return(sum(multiples))
}

t <- Sys.time()
get_sum_multiples_below3(1000000,3,5)
Sys.time() - t


```

</div>


## Exercice 2

By listing the first six prime numbers: 2, 3, 5, 7, 11, and 13, we can see that the 6th prime is 13.

What is the 10 001st prime number?

<div class = "r">
```{r,collapse=T}


nth_prime <- function(nth){
  primes = c(1,2)
  i = 2
  while(length(primes)<nth){
    i <- i+1
    is_prime = TRUE
    for(j in 2:(round(sqrt(i))+1)){
      if(i%%j == 0){
        is_prime = FALSE
      }
      if(is_prime == FALSE){
        break
      }
    }
    if(is_prime == TRUE){
      primes = c(primes,i)
    }
  }
  
  return(primes[nth])
}

t <- Sys.time()
nth_prime(10001)
Sys.time() - t


nth_prime2 <- function(nth){
  primes = c(2,3)
  i = 3
  while(length(primes)<nth){
    i <- i+1
    ratio = i%%seq(2,round(sqrt(i))+1) == 0
    if(!any(ratio)){
      primes = c(primes,i)
    }
  }
  return(primes[nth])
}


t <- Sys.time()
nth_prime2(10001)
Sys.time() - t


```

</div>


# Exercice 3

2520 is the smallest number that can be divided by each of the numbers from 1 to 10 without any remainder.

What is the smallest positive number that is evenly divisible by all of the numbers from 1 to 20?

<div class = "r">
```{r}
# R


smalles_multiple <- function(n){
  dividers = seq(1,n)
  i = 10
  is_multiple = FALSE
  while(is_multiple == FALSE){
    modulos = i%%dividers
    len_multiple = length(modulos[which(modulos == 0)])
    if(len_multiple == length(dividers)){
      is_multiple = TRUE
    } else {
      i = i + 1
    }
  }
  print(i)
}

t = Sys.time()
smalles_multiple(10)
Sys.time() - t

t = Sys.time()
#smalles_multiple(20)
Sys.time() - t



smalles_multiple <- function(n){
  dividers_all = seq(1,n)
  multiple = 1
  i = 10
  for(j in seq(1,n,10)){
    dividers = dividers_all[j:(j+9)]
    is_multiple = FALSE
    while(is_multiple == FALSE){
      modulos = i%%dividers
      len_multiple = length(modulos[which(modulos == 0)])
      if(len_multiple == length(dividers)){
        is_multiple = TRUE
        multiple = i
      } else {
        i = i + multiple
      }
    }
  }
  print(i)
}

t = Sys.time()
smalles_multiple(10)
Sys.time() - t


t = Sys.time()
smalles_multiple(20)
Sys.time() - t


```


</div>

# Exercice 4

$2^{15}$=32768 and the sum of its digits is 3 + 2 + 7 + 6 + 8 = 26.

What is the sum of the digits of the number $2^{1000}$?

<div class = "r">
```{r}
library(gmp)
bigq = as.bigz(2**1000)
sum(as.integer(stringr::str_split(as.character(bigq),'')[[1]]))
```
</div>


# Exercice 5
 
A palindromic number reads the same both ways. The largest palindrome made from the product of two 2-digit numbers is 9009 = 91 × 99.

Find the largest palindrome made from the product of two 3-digit numbers.


<div class = "r">
```{r}

vec_palindrome = function(seq){
  vec1 = rep(seq,each = length(seq))
  vec2 = rep(seq,length(seq))
  product = vec1*vec2
  
  product_inv = stringi::stri_reverse(product)
  palindromes = product[which(product ==  product_inv)]
  return(max(palindromes))
}

t = Sys.time()

three_digits = 100:999
vec_palindrome(three_digits)

Sys.time() - t

```
</div>
<a href="3_Arrays_Vectors.html"> Go back to chapter 2</a>
